Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-25 paper.tex
writing

When the first quantum physicists realised that the position of a particle is essentially uncertain, they were destined to seek the conditions that cause it to follow a normal law of error, and discover the gaussian wave packet.  They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.  These problems were also solved by the gaussian wave packet.  With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.

@@

In the 1970s, Heller put the two ideas together, proposing to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.  The experience of quantum optics shows that many systems can be approximated by a reasonable number of components, and variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.

@@

The expansion introduced by Glauber\cite{prx-131-2766} is performed by the operator
$$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.  The function $f$ is a function of a complex variable, such that $e^{-½|α|²}f(α)$ is square-integrable.  Bargmann\cite{} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.  In this paper, the application of a linear function is written as a product, and functions whose range is ket space are denoted by kets, usually with a greek letter.  So $\G f$ is the ket that results from applying the linear operator $\G$ to the phase space distribution $f$.

@@

Coherent states are not orthogonal, but their brackets follow the law
$$〈β|α〉=e^{-½|β|²-½|α|²+β*α},$$
whence
$$|〈β|α〉|²=e^{-|β-α|²}.$$
The projector satisfies the law
$$\int |α〉〈α|\,d²α=π.$$
For an orthonormal basis, the right hand side would be 1 instead of $π$.  It follows that the operator $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
$$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.  Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.  Further properties of the coherent states will be stated as required.  Glauber's summary of them is paraphrased in every quantum optics textbook ever written.

Prior to Glauber, Bargmann pointed out that the Hilbert space of wavefunctions is isomorphic to the space of entire functions of a complex variable, under the inner product
$$\langle g,f\rangle=\int g*(z)f(z)\,d²z.$$
He introduced the isometry
$$\langle Γ^+|ψ\rangle(α)=π^{-½}\langle α|ψ\rangle,$$
which Glauber showed was a pseudoinverse of $|Γ\rangle$, in the sense that
$$|Γ\rangle\langle Γ^+|=1.$$
The other combination, $\langle Γ^+|Γ\rangle$, is an orthogonal projector from the functions of a complex variable onto the entire functions, under the Bargmann inner product.  It follows that $f=\langle Γ^+|ψ\rangle$ is the function with least norm that satisfies $|Γ\rangle f=|ψ\rangle$.  As shown above, $‖f‖=‖ |ψ\rangle‖$.

If, following Bargmann, we identify the quantum space with the entire functions, $|Γ\rangle$ becomes the projector $\langle Γ^+|Γ\rangle$.  Therefore the singularity of $|Γ\rangle$ is easily resolved.  If we constrain the norm of $f$ to be not much greater than the norm of $|ψ\rangle$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $\langle Γ^+|ψ\rangle$.  And $\langle Γ^+|$, an isometry, is as nice a function as could possibly be desired.  So it is curious that the discrete version of this problem causes so much trouble.

@@

Hypothesis: The singular values of $|Γ〉$ are 0 and 1, and the space belonging to 1 comprises the entire functions.

There are two ways one could set about proving the Hypothesis.  Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.  Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.

This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.  My numerical experiment suggests otherwise; let's ignore that possibility for now.

Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
$$f(z)=∑c_n{zⁿ\over (n!)^½},$$
for some normalised vector $c$.

  Our expansion function is 
$$g(α)=f(α)e^{-½|α|²}.$$
If $f$ has the Taylor expansion
$$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
then 
$$Gg = ∑_n c_n|n〉.$$
The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
$$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
the Glauber preimage of $|n〉$.  We have 
$$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
so the squared norm is
$$‖g‖²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
According to Wolfram,
$$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
so
$$‖g‖=π^½.$$

Assuming that there is no way to expand a state with a smaller norm than an entire function, this does it.  Or at least I think so.  I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.

@@

If the domain and range of an operator are the same, and the operator has singular values 0 and 1, then that operator is an orthogonal projector onto its range.  The operator $|Γ〉$ is a generalisation of that.  It is the composition of an orthogonal projection of function space onto the entire functions followed by an isometric mapping of the entire functions onto the kets.  The orthogonality and isometry are under the Bargmann inner product for phase space functions.  Therefore, if the amplitudes of the coherent states are held constant, and the discretised problem is solved by adjusting their weights, the solutions that come from an SVD should be very stable indeed.  This might depend somewhat on how evenly the amplitudes are spaced: I could start with a regular grid, and add Gaussian noise to see what changed.

So $G$ is a projection: if the inner product is normalised properly, $G$ keeps the component of $f$ in the space of entire functions, and discards the orthogonal component.  The continuous problem is simply underdetermined, and the inverse problem can be regularised very easily by excluding components with zero singular values from the solution.  That isn't what I see with discrete versions: they seem to have a steady decay of singular values, not a sudden drop to zero.  This must be entirely an artefact of the discretisation.

@@

Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.  The obvious choice is $g_r(α*)=(〈α_r|G^+)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.  The condition becomes
$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
	& & \ddots & \cr
	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
= \pmatrix{π\cr π\cr\vdots\cr π}.
$$
Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.  It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.  The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.  This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.

>>>>

When they realised that positions and momenta are essentially uncertain, the first quantum physicists were bound to discover the gaussian wave packet by seeking conditions that produce a normal distribution of probability.  They soon understood the free particle and turned to the harmonic oscillator, investigating the form of its ground state and the way its excited states generate sinusoidal motion in the classical limit.  These problems were also solved by the gaussian wave packet.  With its width fixed by equipartition of energy and the uncertainty principle, it became the coherent state.

@@

In the 1970s, Heller put the two ideas together, proposing to simulate quantum dynamics variationally, using finite superpositions of coherent states as the approximants.  This is an appealing approach.  Every quantum state can be approximated arbitrarily well as the number of components in the superposition increases.  The experience of quantum optics shows that, in many systems, the approximation converges with a reasonable number of components.  Also, variational dynamics should avoid the covergence problems that limit the Monte-Carlo methods.

@@

Expansion over coherent states is a linear operator from functions of a complex variable to quantum states,
$$\G f=π^{-½}\int |α〉e^{-½|α|²}f(α*)\,d²α.$$
The notation $\int f(α)\,d²α$ means $\int_{-∞}^{∞}\int_{-∞}^{∞} f(x+iy)\,dx\,dy$.  The function $f$ is such that $e^{-½|α|²}f(α)$ is square-integrable.  Bargmann\cite{pam-14-187} showed that these functions form a Hilbert space under the inner product $〈g,f〉=\int e^{-|α|²}g*(α)f(α)\,d²α$; this paper refers to it as the space of phase space distributions, or just distributions.  When norms of functions are written below, the norm is the one derived from this inner product.  In this paper, $\G f$ denotes the ket that results from applying the linear operator $\G$ to the phase space distribution $f$; application and composition of linear operators is shown by juxtaposition, and functions whose values are quantum states are written as kets, usually with a greek letter.

@@

Coherent states are not orthogonal, but their brackets follow the law
$$〈β|α〉=e^{-½|β|²-½|α|²+β*α},$$
whence
$$\bigl|〈β|α〉\bigr|²=e^{-|β-α|²}.$$
The projector satisfies
$$\int |α〉〈α|\,d²α=π.$$
This is similar to the law for an orthonormal basis, except that the right hand side is $π$ instead of 1.  Further properties of the coherent states will be stated as required.  Glauber's summary of them\cite{prx-131-2766} is paraphrased in every quantum optics textbook ever written.

% It follows that $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
% $$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
% Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.  Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.

Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
$$〈 Γ⁺|ψ〉(α)=π^{-½}〈 α|ψ〉.$$
Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.  It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.  The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.

If, following Bargmann, we identify the quantum space with the entire functions, $|Γ\rangle$ becomes the projector $\langle Γ⁺|Γ\rangle$.  Therefore the singularity of $|Γ\rangle$ is easily resolved.  If we constrain the norm of $f$ to be not much greater than the norm of $|ψ\rangle$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $\langle Γ⁺|ψ\rangle$.  And $\langle Γ⁺|$, an isometry, is as nice a function as could possibly be desired.  So it is curious that the discrete version of this problem causes so much trouble.

@@

% Hypothesis: The singular values of $|Γ〉$ are 0 and 1, and the space belonging to 1 comprises the entire functions.
% 
% There are two ways one could set about proving the Hypothesis.  Firstly, one could show that any function orthogonal to the Gaussian analytics annihilates $G$.  Otherwise, one could show that the norms of all Gaussian analytics are scaled by the same factor going to F
% ock space, and that they are a pre-image of Fock space.  The second approach sounds simpler.
% 
% This won't be a rigourous argument: it could turn out that there's another class of expansions with half the norm, and the entire functions have components belonging to the 0 singular value.  My numerical experiment suggests otherwise; let's ignore that possibility for now.
% 
% Let $f$ be a normalised entire function, in Glauber's sense.  I.e., 
% $$f(z)=∑c_n{zⁿ\over (n!)^½},$$
% for some normalised vector $c$.
% 
%   Our expansion function is 
% $$g(α)=f(α)e^{-½|α|²}.$$
% If $f$ has the Taylor expansion
% $$f(z)=∑_n c_n{zⁿ\over(n!)^½},$$
% then 
% $$Gg = ∑_n c_n|n〉.$$
% The functions $zⁿ$ for $n≥0$ span the entire functions (or their span is a dense subset, or something like that), so let's find the norm of 
% $$g(α)=e^{-½|α|²}{αⁿ\over(n!)^½},$$
% the Glauber preimage of $|n〉$.  We have 
% $$|g(α)|²={1\over n!}e^{-|α|²}|α|^{2n},$$
% so the squared norm is
% $$‖g‖²={1\over n!}\int ₀^∞2πr^{2n+1}e^{-r²}\,dr={2π\over n!}\int ₀^∞r^{2n+1}e^{-r²}\,dr.$$
% According to Wolfram,
% $$\int ₀^∞r^{2n+1}e^{-r²}\,dr={n!\over 2},$$
% so
% $$‖g‖=π^½.$$
% 
% Assuming that there is no way to expand a state with a smaller norm than an entire function, this does it.  Or at least I think so.  I should look up exactly what the uniqueness properties are for singular value expansions in infinite dimensional spaces.

@@

@@

Since there are only $R$ weights to adjust, the condition $|\vec g|=|g|$ can be satisfied only for $R$ linearly independent functions $g_r$.  The obvious choice is $g_r(α*)=(〈α_r|G⁺)(α*)=〈α|α_r〉=\exp(-½|α|²-½|α_r|²+α*α_r)$, with $|\hat g_r^s|²=w_s²\exp(-|α_s-α_r²)$, and $|g_r|²=π$.  The condition becomes
$$\pmatrix{1 & e^{-|α₁-α₂|²} &  & e^{-|α₁-α_R|²} \cr
	e^{-|α₂-α₁|²} & 1 &  & e^{-|α₂-α_R|²} \cr
	& & \ddots & \cr
	e^{-|α_R-α₁|²} & e^{-|α_R-α₂|²} & & 1}
\pmatrix{w₁²\cr w₂²\cr\vdots\cr w_R²}
= \pmatrix{π\cr π\cr\vdots\cr π}.
$$
Unfortunately, for random samples $A$, this tends not to have a solution: some of the $w_r²$ would have to be negative.  It looks like the amplitudes on the edges of the grid need large weights, because half of their corresponding gaussian lies outside the grid, and isn't sampled.  The gaussians centred inside the grid are sampled at all these points, and need negative weights at the inside points to compensate.  This sounds like the problem that Chebyshev points solve, where the function is sampled more closely at the edges of the domain.
